### 2022-08-05 front/declaration.tex
intro/ch2 edits

This thesis contains work taken from the publication \cite{blnos2022} which approximates the operator analytic expressions of \cite{bo2014} via the DG method. I am a co-author on this paper. The main conceptualisation and initial writing of the manuscript was done by Vikram Sunkara, Giang Nguyen and Nigel Bean. I extended the manuscript in the following ways.
\begin{itemize}
    \item Significant additions to Section~2 to introduce the partition induced by the approximation scheme to the theoretical operators of \cite{bo2014} to help elucidate the connection between the theoretical operators and the approximation counterparts. 
    \item The addition of Section~4.4 which introduces exact boundary dynamics. Prior to this boundary conditions were approximated by a small interval in the DG scheme (in this thesis, I generalised the boundary conditions further).
    \item Added and proved Corollary~4.1.
    \item Formally deriving, and describing, the approximation to the operator \(\mathbb R\) in Section~5.1. 
    \item Introduced the concept of operator kernels to the paper (used throughout Section~5.5).
    \item Wrote the final version of the code (Vikram did the initial version).
    \item Extended the numerical experiments to consider higher-order bases, smaller cell-widths, elucidating issues around truncation errors, and computing confidence intervals. 
    \item Added Appendix~A on the properties of the DG approximation to the generator of the fluid queue.
    \item Added details to Appendix~B. 
    \item I also made significant contributions to the writing (and rewriting) the manuscript, submission, and responding to reviewers comments. 
\end{itemize}
To clarify which parts of this thesis are related to the paper \cite{blnos2022}, the following sections were taken, almost verbatim except for minor editing, notational changes so that the notation matches that of the thesis, and the addition more general boundary conditions,
\begin{itemize}
    \item Section~\ref{sec: ffq intro},
    \item Chapter~\ref{ch:galerkin}, except for Sections~\ref{sec: other applications}-\ref{sec: slope limiting},
    \item Appendix~\ref{appendix:example},
    \item Appendix~\ref{sec:properties}.
\end{itemize}

>>>>

This thesis contains work taken from the publication \cite{blnos2022} which approximates the operator analytic expressions of \cite{bo2014} via the DG method. I am a co-author on this paper. The main conceptualisation and initial writing of the manuscript was done by Vikram Sunkara, Giang Nguyen and Nigel Bean. I extended the manuscript in the following ways.
\begin{itemize}
    \item Significant additions to Section~2 to introduce the partition induced by the approximation scheme to the theoretical operators of \cite{bo2014} to help elucidate the connection between the theoretical operators and the approximation counterparts. 
    \item The addition of Section~4.4 which introduces exact boundary dynamics. Prior to this boundary conditions were approximated by a small interval in the DG scheme (in this thesis, I generalised the boundary conditions further).
    \item The addition and proof of Corollary~4.1.
    \item The formal definition and derivation of the approximation to the operator \(\mathbb R\) in Section~5.1. 
    \item Introduction of the concept of operator kernels to the paper (used throughout Section~5.5).
    \item The final version of the code (Vikram did the initial version).
    \item Extension of the numerical experiments to consider higher-order bases, smaller cell-widths, elucidating issues around truncation errors, and computing confidence intervals. 
    \item Addition of Appendix~A on the properties of the DG approximation to the generator of the fluid queue.
    \item Additional details in Appendix~B. 
    \item I also made significant contributions to the writing (and rewriting) the manuscript, submission, and responding to reviewers' comments. 
\end{itemize}
To clarify which parts of this thesis are related to the paper \cite{blnos2022}, the following sections were taken, almost verbatim except for minor editing, notational changes so that the notation matches that of the thesis, and the addition more general boundary conditions,
\begin{itemize}
    \item Section~\ref{sec: ffq intro},
    \item Chapter~\ref{ch:galerkin}, except for Sections~\ref{sec: other applications} and \ref{sec: limiting and linearity},
    \item Appendix~\ref{appendix:example},
    \item Appendix~\ref{sec:properties}.
\end{itemize}

### 2022-08-05 intro/chapter.tex
intro/ch2 edits

Typically, the phase, \(\varphi(t)\), represents the underlying operating state or environment of a system at time \(t\). The level variable, \(X(t)\), describes some continuous variable of the system. Stochastic fluid queues have found a variety of applications such as telecommunications (see \cite{anick1982}, a canonical application in this area), power systems \citep{hydro}, risk processes \citep{betal2005}, environmental modelling \citep{wurm2020}, dam storage \citep{loynes1962}, and more. For example, \cite{anick1982} model a \emph{data handling switch} in which there are \(N\) information sources which asynchronously switch from ON to OFF, or OFF to ON. The phase process \(\varphi(t)\) models the number of data sources which are ON at time \(t\), and the rate at which data is received when the buffer is on is assumed (without loss of generality) to be 1. The switch stores data in a buffer and processes the data at rate \(c\), thus the net rate of change of the buffer at time \(t\) when there are \(r\) information sources which are ON is \(r-c\). The level process, \(X(t)\), models the amount of data in the buffer at time \(t\) and the rate at which \(X(t)\) changes which \(\varphi(t)=r\) is \(c_r=r-c\). The model described above could be used to aid in the analysis and design of the data handling switch, for example, by determining the minimum rate \(r\) for which the system is \emph{stable} in steady state, which practically means that the processing power of the switch is large enough that the amount of data in the buffer does not grow without bound.

@@

The addition of the second level process \(\{Y(t)\}\) greatly increases the flexibility and modelling potential. However, the added flexibility comes at a cost to the in the analysis of the model. For both fluid-fluid queue and classical fluid queues, performance measures such has the stationary distribution, and the Laplace-transforms of hitting times and return times, can be derived in terms of the infinitesimal generator of the modulating process. For fluid queues the modulating process is a CTMC for which the infinitesimal generator is a matrix. Thus, expressions for the performance measures of fluid queues are matrix expressions which are readily computable. For fluid-fluid queue the modulating processes is a fluid queue for which the infinitesimal generator is a matrix of differential operators. Thus, expressions for the performance measures of fluid-fluid queues are matrices of differential operators which, in all but the simplest of cases, are not readily computable.

The analysis of \cite{bo2014} is in principle similar to the matrix-analytic methods of \cite{bean2005}, and derives results about the second level process \(\{Y(t)\}_{t\geq0}\) in terms of the infinitesimal generator (a differential operator) of the fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a discretisation of the infinitesimal generator of the fluid queue can be used. To this end, to date, two possible discretisations have been suggested. Taking a differential equations-based approach, \cite{blnos2022} use the discontinuous Galerkin (DG) method to discretise the operator, while \cite{bo2013} take a stochastic modelling and matrix-analytic methods approach to approximate the fluid-queue by a quasi-birth-and-death (QBD) process. The QBD approximation of \cite{bo2013} is derived via a uniformisation argument, so we refer to it as the uniformisation scheme throughout this thesis. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the uniformisation scheme of \cite{bo2013} is a subclass of the former; the uniformisation scheme can be viewed as the simplest DG scheme where cells are all have equal width, the operator is projected onto a basis of piecewise constant functions, and an upwind flux is used to approximate the flow of mass between cells. This is also equivalent to a finite-volume scheme.

@@

The structure of this thesis is as follows. The next chapter, Chapter~\ref{ch:math}, is dedicated to mathematical preliminaries and introduces the main mathematical objects and tools which we will need. In particular, we go into detail describing operators arising in the analysis of fluid-fluid queues and introducing them in such a way that makes clear exactly how the approximations we develop correspond to the theoretical operators. Chapter~\ref{ch:math} also introduces the existing literature and gives context to this thesis.

Chapter~\ref{ch:galerkin} demonstrates a way that we can use the discontinuous Galerkin (DG) method to approximate certain operators and distributions of fluid and fluid-fluid queues. Due to issues with oscillations in the approximations produced by the DG scheme, at the end of Chapter~\ref{ch:galerkin} we also describe \emph{slope limiting} -- a method which can be used to prevent oscillations and negativity. At the end of Chapter~\ref{ch:galerkin} we also introduce other numerical procedures which we use in our numerical investigations in Chapter~\ref{sec: numerics}.

In the following chapter, Chapter~\ref{sec: construction and modelling}, we develop a new QBD-RAP approximation scheme which, due to its interpretation as a stochastic process, ensures all estimates of probabilities will be non-negative. The chapter takes a stochastic modelling approach to developing the approximation scheme.

Once we have established the new approximation scheme, we then prove a convergence of the approximation scheme in Chapters~\ref{sec: conv}~and~\ref{ch: global results} via certain Laplace transforms with respect to time. Chapter~\ref{sec: conv} utilises methods and ideas relating specifically to QBD-RAPs and matrix exponential distributions. Ultimately, Chapter~\ref{sec: conv} proves that, on the event that the QBD-RAP has not-yet seen an \emph{orbit restart epoch} (which is much like a change of level, but not always), certain Laplace transforms with respect to time of the QBD-RAP scheme converge to corresponding Laplace transforms with respect to time of the fluid queue. This is a local convergence result as it relates only to convergence to the fluid queue in an interval.

Chapter~\ref{ch: global results} then looks to extend the local convergence results of Chapter~\ref{sec: conv} to a global convergence result on the whole domain of approximation. Chapter~\ref{ch: global results} uses more traditional Markov process arguments which rely on things such as the strong Markov property, time-homogeneity and the Law of total probability. In Chapter~\ref{ch: global results} we first consider the discrete-time process which the process embedded in the QBD-RAP at times when the QBD-RAP sees an orbit restart epoch. We prove that the embedded process converges in distribution to a corresponding embedded process of the fluid queue. The rest of Chapter~\ref{ch: global results} then attempt to prove a global converge result. The main result being Theorem~\ref{thm: big thm} which states that the QBD-RAP approximation scheme converges weakly (with respect to the spatial and temporal variable) to the distribution of the fluid queue.

@@

To help the reader understand the notation used in the DG scheme, we provide a small toy model example in Appendix~\ref{appendix:example}. Appendix~\ref{sec:properties} provides a proof that the DG scheme conserves probability. The rest of the appendices contain technical results relating to proving the convergence of the QBD-RAP approximation scheme. Appendix~\ref{appendix: sec: 2} proves that the \emph{closing operators} introduced as part of the QBD-RAP scheme have the properties we claim they do to prove convergence in Chapter~\ref{sec: conv}. Appendix~\ref{app:extend conv} extends some results from Chapter~\ref{sec: conv} to a setting which requires slightly less computation. Lastly, Appendix~\ref{appendix: kronecker} provides some algebraic results which help us to manipulate certain Laplace transforms from Chapter~\ref{sec: conv}.

>>>>

Typically, the phase, \(\varphi(t)\), represents the underlying operating state or environment of a system at time \(t\). The level variable, \(X(t)\), describes some continuous variable of the system. Stochastic fluid queues have found a variety of applications such as telecommunications (see \cite{anick1982}, a canonical application in this area), power systems \citep{hydro}, risk processes \citep{betal2005}, environmental modelling \citep{wurm2020}, dam storage \citep{loynes1962}, and more. For example, \cite{anick1982} model a \emph{data handling switch} in which there are \(N\) information sources which asynchronously switch from ON to OFF, or OFF to ON. The phase process \(\varphi(t)\) models the number of data sources which are ON at time \(t\), and the rate at which data is received from a single source when the buffer is on is assumed (without loss of generality) to be 1. The switch stores data in a buffer and processes the data at rate \(c\), thus the net rate of change of the buffer at time \(t\) when there are \(r\) information sources which are ON is \(r-c\). The level process, \(X(t)\), models the amount of data in the buffer at time \(t\) and the rate at which \(X(t)\) changes when \(\varphi(t)=r\) is \(c_r=(r-c)\times 1(X(t)>0)\). The model described above could be used to aid in the analysis and design of the data handling switch, for example, by determining the minimum rate \(r\) for which the system is \emph{stable} in steady state, which practically means that the processing power of the switch is large enough that the amount of data in the buffer does not grow without bound.

@@

The addition of the second level process \(\{Y(t)\}\) greatly increases the flexibility and modelling potential. However, the added flexibility comes at a cost to the in the analysis of the model. For both fluid-fluid queue and classical fluid queues, performance measures such as the stationary distribution, and the Laplace-transforms of hitting times and return times, can be derived in terms of the infinitesimal generator of the modulating process. For fluid queues the modulating process is a CTMC for which the infinitesimal generator is a matrix. Thus, expressions for the performance measures of fluid queues are matrix expressions which are readily computable. For fluid-fluid queue the modulating processes is a fluid queue for which the infinitesimal generator is a matrix of differential operators. Thus, expressions for the performance measures of fluid-fluid queues are matrices of differential operators which, in all but the simplest of cases, are not readily computable.

The analysis of \cite{bo2014} is in principle similar to the matrix-analytic methods of \cite{bean2005}, and derives results about the second level process \(\{Y(t)\}_{t\geq0}\) in terms of the generator (a differential operator) of the fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a discretisation of the infinitesimal generator of the fluid queue can be used. To this end, to date, two possible discretisations have been suggested. Taking a differential equations-based approach, \cite{blnos2022} use the discontinuous Galerkin (DG) method to discretise the operator, while \cite{bo2013} take a stochastic modelling and matrix-analytic methods approach to approximate the fluid-queue by a quasi-birth-and-death (QBD) process. The QBD approximation of \cite{bo2013} is derived via a uniformisation argument, so we refer to it as the uniformisation scheme throughout this thesis. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the uniformisation scheme of \cite{bo2013} is a subclass of the former; the uniformisation scheme can be viewed as the simplest DG scheme where cells are all have equal width, the operator is projected onto a basis of piecewise constant functions, and an upwind flux is used to approximate the flow of mass between cells. This is also equivalent to a finite-volume scheme.

@@

The structure of this thesis is as follows. The next chapter, Chapter~\ref{ch:math}, is dedicated to mathematical preliminaries and introduces the main mathematical objects and tools which we will need. In particular, we go into detail describing operators arising in the analysis of fluid-fluid queues and introducing them in such a way that makes clear exactly how the approximations we develop correspond to the theoretical operators. Chapter~\ref{ch:math} also introduces the discontinuous Galerkin (DG) method for approximating solutions to differential equations. Due to issues with oscillations in the approximations produced by the DG scheme, at the end of Chapter~\ref{ch:math} we also describe \emph{slope limiting} -- a method which can be used to prevent oscillations and negativity. Chapter~\ref{ch:math} also introduces the existing literature and gives context to this thesis.

Chapter~\ref{ch:galerkin} demonstrates a way that we can use the discontinuous Galerkin method to approximate certain operators and distributions of fluid and fluid-fluid queues. The DG method provides an effective tool for approximating performance measures of fluid-fluid queues. However, the DG method can exhibit oscillations in the presence of discontinuities and these oscillations can cause approximations to certain probabilities to be negative. This motivates the next chapter.

In the next chapter, Chapter~\ref{sec: construction and modelling}, we develop a new QBD-RAP approximation scheme which, due to its interpretation as a stochastic process, ensures all estimates of probabilities will be non-negative. The chapter takes a stochastic modelling approach to developing the approximation scheme.

Once we have established the new approximation scheme, we then prove its convergence in Chapters~\ref{sec: conv}~and~\ref{ch: global results}. Ultimately, Chapter~\ref{sec: conv} proves that, on the event that the QBD-RAP has not-yet seen an \emph{orbit restart epoch} (which is much like a change of level, but not always), certain Laplace transforms with respect to time of the QBD-RAP scheme converge to corresponding Laplace transforms with respect to time of the fluid queue. This is a local convergence result as it relates only to convergence to the fluid queue in an interval.

Chapter~\ref{ch: global results} then looks to extend the local convergence results of Chapter~\ref{sec: conv} to a global convergence result on the whole domain of approximation. Chapter~\ref{ch: global results} uses more traditional Markov process arguments which rely on properties such as the strong Markov property, time-homogeneity and the law of total probability. In Chapter~\ref{ch: global results} we first consider the discrete-time process which is the process embedded in the QBD-RAP at times when the QBD-RAP sees an orbit restart epoch. We prove that the embedded process converges in distribution to a corresponding embedded process of the fluid queue. The rest of Chapter~\ref{ch: global results} proves a global convergence result. The main result is Theorem~\ref{thm: big thm} which states that the QBD-RAP approximation scheme converges weakly (with respect to the spatial and temporal variable) to the distribution of the fluid queue.

@@

To help the reader understand the notation used in the DG scheme, we provide a small toy model example in Appendix~\ref{appendix:example}. Appendix~\ref{sec:properties} provides a proof that the DG scheme conserves probability. The rest of the appendices contain technical results relating to proving the convergence of the QBD-RAP approximation scheme. Appendix~\ref{appendix: sec: 2} proves that the \emph{closing operators} introduced as part of the QBD-RAP scheme have the properties we claim in proving convergence in Chapter~\ref{sec: conv}. Appendix~\ref{app:extend conv} extends some results from Chapter~\ref{sec: conv} to a setting which requires slightly less computation. Lastly, Appendix~\ref{appendix: kronecker} provides some algebraic results which help us to manipulate certain Laplace transforms from Chapter~\ref{sec: conv}.
